Highly Available Transactions: Virtues and Limitations

### 2013-03-26 writeup/conclusion.tex
Another pass at writing

VIII. Conclusion (.5pg)

HATS represent a new class of database system:
* Can't get serializability, but can get other properties
* Key challenge: designing distributed algorithms and architectures for partial failure, asynchrony
* Rearchitecture of DBMSs and better understanding of applications are key

References (1pg)

>>>>

In this paper, we have begun to understand the relationship between
multi-operation, multi-object transactions and high availability. We
have outlined requirements for high availability in a transactional
context, or Highly Available Transactions (HATs) and taxonomized
guarantees spanning both distributed systems and ACID database
literature accordingly. Many transactional guarantees are achievable
with high availability, while a few, like preventing Lost Update and
Write Skew, are not. To understand the benefits of high availability
and low latency, we classified the prevalence and cost of partitions
and network latency both in the literature and via experiment, while
analyzing existing transactions to determine whether HATs were
sufficient to ensure application consistency.

Fundamentally, high availability represents a new constraint for
database designs. While the CAP Theorem has had a large impact on
distributed data storage design, our knowledge of database systems has
not kept pace. The key distributed systems challenge of designing
distributed algorithms and architectures for partial failure and
network asynchrony is largely unaddressed in the literature on
transaction processing systems. We believe that, in order to fully
realize the potential of HAT systems, further rearchitecture and study
of existing database designs as well as better understanding of
application semantics and their relationship to distributed
consistency will be necessary.

### 2013-03-26 writeup/evaluation.tex
Another pass at writing

With a firm understanding of which of several ACID and distributed
consistency properties are available and not, we perform analysis of
their implications for existing algorithms, applications, and
deployments. Specifically:

@@

We begin by considering several existing algorithms that provide
serializability but are not highly available. In summary, all
algorithms in this class require at least one round trip time (RTT)
before committing.

\begin{itemize}
\item \textbf{Two-phase locking} (ignoring deadlocks) requires $T$
  lock operations and 1 unlock operation for a transaction of length
  $T$. If all items are located on a single master server, this
  requires one RTT (for requests originating from servers other than
  the master). If items are located on $S$ ($S \leq T$) different
  servers, two-phase locking requires, at minimum, $2(S-1)$ RTTs---one
  for lock and another for unlock.

\item \textbf{Optimistic concurrency control} allows reads from any
  server but transactions must be checked by a scheduler before
  commit. With one scheduler per database, this requires one RTT
  specifically for concurrency control. If schedulers are sharded by
  database partition, a multi-partition transaction may require
  contacting several different schedulers and, possibly, an atomic
  commit protocol, which becomes substantially more expensive.

\item \textbf{Deterministic transaction scheduling} is ffectively the
  inverse of optimistic concurrency control: a scheduler first orders
  the transactions, then sends them to database replicas for in-order
  execution. These techniques, pioneered in the context of totally
  ordered atomic broadcast systems, have recently seen increased
  attention and effectively require communication overhead similar to
  optimistic concurrency control.

\item\textbf{Multiversion concurrency control} allows readers to read
  older versions of data items but, like OCC, must eventually be
  checked for write-write conflicts, requiring a scheduler.

\end{itemize}

Perhaps unsurprisingly, the reliance on a glocal total order
necessitates a minimum of one round-trip to a designated master or
coordination service for each of these classic algorithms. This is
simply a cost of providing serializability, and the cost of the
(minimum one) round trip will be determined by the deployment
environment, as we saw in Section~\ref{sec:motivation}.

There also are several mechanisms that do not provide serializability
but provide transactional semantics with varying availability
properties:

\begin{itemize}
\item \textbf{Lock-based protocols} such as those used to provide weak
  isolation in Gray's original paper require at least one RTT to the a
  lock server (or, alternatively, a majority quorum). In the presence
  of failure, distributed locking is challenging.
\item The \textbf{MDCC} commit protocol uses a Paxos variant and escrow
  transactions to provide Repeatable Read isolation with 1-2 RTTs. (TODO: CHECK)
\item \textbf{Parallel Snapshot Isolation} as developed by the Walter
  system provides (unavailable) snapshot isolation variant and, even
  when writes commute, requires 1 RTT to enforce read
  recency. However, PSI without recency requirements could be made
  highly available.
\item \textbf{Eiger} provides read-only and write-only transactions
  with item cut isolation and causal consistency. It relies on sticky
  high availability and groups within a datacenter with linearizable
  clusters. In the presence of server failure during write
  transactions, reads and writes may become unavailable. \textbf{COPS}
  provides similar semantics for read-only transactions but does not
  suffer from unavailability as it does not have a coordinator.
\item \textbf{Chan and Gray's Read-only transactions} use a similar
  algorithm to Eiger to provide read-only transactions with item cut
  isolation and causal consistency but coordinate on reads instead of
  writes.
\item \textbf{Bolt-on Causal Consistency} and \textbf{Swift} can
  provide read-only and write-only transactions with item cut
  isolation and causal consistency with sticky high availability and
  both rely on client-side caching.
\item \textbf{Bayou} provides several session guarantees but, without
  client-side caching, actually faces unavailability in the presence
  of failure. (TODO: CHECK)
\item \textbf{Brantner's S3 Database} uses caching to provide several
  guarantees such as XX and YY but is unavailable during failures
  (TODO: CHECK).
\end{itemize}

@@

Recent work on the CALM Theorem, Lattice-based state update, and
Commutative and Replicated Data Types demonstrates that, if updates
logically commute, then they can be performed in different orders at
different locations in a distributed system. Accordingly, as long as
all writes are delivered to all replicas, then a system executing
monotonic logic with commutative operators should not suffer from
application-level consistency anomalies as a result of Lost Update or
Write Skew anomalies. However, applications with non-monotonic state
mutation will not, in general, be able to maintain application-level
consistency constraints. Similarly, applications requiring
write visibility between clients should opt for unavailability instead
of possibly missing a write visibility deadline.

As an example of an actual system rearchitecture,
less than one year after its announcement, Yahoo!'s PNUTS developers
explicitly added a support for weaker, highly available operation,
explaining that ``strict adherence [to strong consistency] leads to
difficult situations under network partitioning or server
failures...in many circumstances, applications need a relaxed
approach''~\cite{pnuts-update}.

Absent a necessary and sufficient result describing exactly which
application logics can be achieved with high availability and to make
these requirements more concrete, we consider an example application:
the TPC-C benchmark. Our goal is \textit{not} to provide a performance
analysis, as any HAT configuration will be non-compliant. Rather, we
examine the five queries contained in the benchmarks---ostensibly
representative of a standard transaction processing application---to
understand which are achievable in a HAT context. TPC-C consists of
five transactions, capturing operation of a real-world warehouse,
including sales, payments, and deliveries. We analyze each transaction
and its interactions below.

\begin{itemize}
\item \textbf{New-Order} places an order for a variable quantity of
  data items, updating warehouse stock as needed. The New-Order
  transaction selects a sales district, assigns the new order an order
  ID number, adjusts the remaining warehouse stock, and writes a
  placeholder entry for the pending order. There are two mutations we
  must be consider in detail.

  First, assigning an order ID number is challenging. We can easily
  assign a \textit{unique} order ID, by, say, concatenating the
  customer ID with a timestamp and server ID. However, the TPC-C
  specification requires that order numbers are assigned sequentially,
  which we cannot guarantee in HATs due to Lost Update. Accordingly,
  HATs can allow placement of new orders with unique identifiers but
  cannot provide compliant TPC-C execution due to the non-monotonic
  counter increase.

  Second, decrementing inventory counts is a potentially risky
  operation as, without coordination, inventory counts may become
  negative. However, TPC-C automatically ``restocks'' inventory counts
  (by 91 items) when there are 10 or fewer items
  remaining. Accordingly, even in the presence of concurrent new
  orders to the same item, the stock will never fall below
  zero. However, an $R$ replica HAT system might end up with $91(R-1)$
  more stock than a single-site implementation.

\item \textbf{Payment} updates running balances for warehouses,
  districts, and customer records as well as providing an audit
  trail. All balance operations can be made commutative, so the
  transaction can execute under HATs.

\item \textbf{Order-Status} is a read-only transaction and can be
  safely executed under HATs. The TPC-C specification does not require
  recency bounds but, in practice, they may be useful. If a client is
  sticky with her replica set, she will read her own writes, which may
  be sufficient.

\item \textbf{Delivery} represents the fulfillment of a pending order
  created by a New Order transaction: it deletes the order from the
  pending list, updates the customer's balance, updates the order's
  carrier ID and delivery time, and updates the customer balance. The
  challenge in delivery comes in deletion, which is non-monotonic: an
  item might be removed twice from the pending table. If delivery can
  be made idempotent---that is, if multiple Delivery transactions have
  the same effect as one---then multiple clients can execute the same
  Deliveries without fatal consequences.

  In practice, given that a physical shipment is likely only delivered
  once, it is unlikely that multiple agents will attempt a Delivery
  transaction for the same order simultaneously. However, in this
  case, we have simply moved the non-monotonicity to the real world:
  the carrier that picks up the package for an order must ensure that
  she is the only carrier who has done so. While we can program around
  the other operations in the order, HATs cannot--without the help of
  business-specific processes--serve Delivery transactions. That said,
  New Orders may be placed.

\item \textbf{Stock-Level} is another read-only transaction and can be
  executed without any non-HAT guarantees, with the same caveats
  regarding recency.
\end{itemize}

Throughout execution, TPC-C requires the maintainence of several
integrity constraints. For example, Consistency Condition 1 (3.3.2.1)
requires that each warehouse's sales count must reflect the sum of its
subordinate sales districts. This integrity constraint spans two
tables but, given the ability to update both tables atomically via
transactional atomicity, can be easily maintained. Consistency
Conditions 4 through 12 (3.3.2.4-12) can similarly be satisfied by
applying updates atomically across tables. Consistency Conditions 2
and 3 (3.3.2.2-3) concern order ID assignment and are more
problematic: effectively, each New Order placed by a customer in a
district must have an ID that is one the prior order in that
district. This requires preventing Lost Update within each district's
table. However, as we have seen, the required transactions can execute
without these two conditions.

In summary, many---but not all---TPC-C transactions are well-served by
HATs. The two problematic transactions---New Order and Payment---rely
on non-monotonic state updates. The former transaction can be modified
to ensure ID uniqueness at the expense of dense ID ordering, while the
latter is an inherently non-monotonic action requiring external
compensation or stronger consistency protocols. We expect that,
especially for read-dominated workloads found in many online services,
HAT guarantees may suffice for many applications. Finally, while TPC-C
is not subject to write skew anomalies, we note that many TPC-E
isolation tests, such as one that requires simultaneously modifying a
product description and its image, are satisfied by HATs.

>>>>

With a firm understanding of which of several ACID and distributed
consistency properties are available and which are not, we begin to
understand their implications for existing algorithms, applications,
and deployments. Specifically:

@@

\vspace{.5em}\noindent\textbf{Serializablity} We begin by considering several
existing algorithms that provide serializability but are not highly
available. In summary, all algorithms in this
class~\cite{bernstein-book} require at least one round trip time (RTT)
before committing.

\vspace{.5em}\noindent\textit{Two-phase locking} (ignoring deadlocks) requires $T$ lock
operations and 1 unlock operation for a transaction of length $T$. If
all items are located on a single master server, this requires one RTT
(for requests originating from servers other than the master). If
items are located on $S$ ($S \leq T$) different servers, two-phase
locking requires, at minimum, $2(S-1)$ RTTs---one for lock and another
for unlock.

\vspace{.5em}\noindent\textit{Optimistic concurrency control} allows
reads from any server but transactions must be checked by a scheduler
before commit. With one scheduler per database, this requires one RTT
specifically for concurrency control. If schedulers are sharded by
database partition, a multi-partition transaction may require
contacting several different schedulers and, possibly, an atomic
commit protocol, which becomes substantially more
expensive.

\vspace{.5em}\noindent\textit{Deterministic transaction scheduling} is
effectively the inverse of optimistic concurrency control: a scheduler
first orders the transactions, then sends them to database replicas
for in-order execution. These techniques, pioneered in the context of
totally ordered atomic broadcast
systems~\cite{deterministic-scheduling}, have recently seen increased
attention~\cite{calvin} and effectively require communication overhead
similar to optimistic concurrency control. Finally,

\vspace{.5em}\noindent\textit{Multiversion concurrency control} allows readers to read older
versions of data items but, like OCC, must eventually be checked for
write-write conflicts, requiring a scheduler.

\vspace{.5em} Perhaps unsurprisingly, the reliance on a global total
order necessitates a minimum of one round-trip to a designated master
or coordination service for each of these classic algorithms. This is
simply a cost of providing serializability, and the cost of the
(minimum one) round trip will be determined by the deployment
environment, as we saw in Section~\ref{sec:motivation}.

\vspace{.5em}\noindent\textbf{Non-serializablity} While it is possible to use
several of the prior mechanisms for HAT systems---for example, a
multi-version storage system is a reasonable candidate for the basis
of a HAT database, albeit with write conflict checking
disabled---several techniques, like locking, are not suitable for
HATs. In fact, there are several mechanisms that do not provide
serializability but provide transactional semantics with varying
availability properties:

\vspace{.5em}\noindent\textit{Lock-based protocols} such as those used to provide weak
  isolation in Gray's original proposal~\cite{gray-isolation} require
  at least one RTT to the a lock server (or, alternatively, a majority
  quorum). In the presence of failure, distributed locking is
  challenging.

\vspace{.5em}\noindent The \textit{MDCC} commit protocol uses a Paxos
variant and escrow transactions to provide read committed (without
Lost Update) and Non-monotonic Snapshot Isolation with 1-2 RTTs but is
unavailable due to Lost Update~\cite{mdcc}.

\vspace{.5em}\noindent\textit{Parallel Snapshot Isolation} as
developed by the Walter system provides (unavailable) snapshot
isolation variant and, even when writes commute, requires 1 RTT to
enforce read recency~\cite{walter}. However, like other Snapshot
Isolation variants, PSI could be made highly available by giving up
recency requirements and Lost Update prevention.

\vspace{.5em}\noindent\textit\textbf{Eiger} provides read-only and
write-only transactions with item cut isolation and causal
consistency. It relies on sticky high availability with a linearizable
set of replicas within a datacenter. This provides low latency across
data centers. In the presence of server failure during write
transactions, reads and writes may become unavailable. Eiger's
predecessor \textit{COPS} provides similar semantics for read-only
transactions but does not suffer from unavailability as it does not
rely have a coordinator.

\vspace{.5em}\noindent\textit{Chan and Gray's Read-only transactions}
use a similar algorithm to Eiger to provide read-only transactions
with item cut isolation and causal consistency but client version
selection coordination is performed on reads instead of
writes~\cite{readonly}.

\vspace{.5em}\noindent\textit{Bolt-on Causal
  Consistency}~\cite{bolton} and \textit{Swift}~\cite{swift} can
provide read-only and write-only transactions with item cut isolation
and causal consistency with sticky high availability and both rely on
client-side caching.

\vspace{.5em}\noindent\textit{Bayou} provides several session
guarantees but, in its original implementation, may face unavailabilty
in the presence of failure~\cite{sessionguarantees}.

\vspace{.5em}\noindent\textit{Brantner's S3 Database} uses caching to
provide several session and transactional guarantees on top of cloud
infrastructure, often relying on linearizable queues and, accordingly,
are often unavailable~\cite{kraska-s3}

\vspace{.5em}

As we have seen, it is possible to implement many guarantees weaker
than serializability---including guarantees that are achievable with
high availability---and not achieve high availability. These
designs---though often unavailable---offer benefits unavailable to
HATs: for example, lock-based read committed, provide recency
guarantees in addition to read committed isolation! We expect that
further understanding the trade-offs between semantic strength,
performance, and storage and metadata requirements will remain an area
of active research for the foreseeable future.

@@

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent work on
the CALM Theorem~\cite{calm}, Lattice-based state
update~\cite{blooml}, and Commutative and Replicated Data
Types~\cite{crdt} demonstrates that, if updates logically commute,
then they can be performed in different orders at different locations
in a distributed system. Accordingly, as long as all writes are
delivered to all replicas, then a system executing monotonic logic
with commutative operators should not suffer from application-level
consistency anomalies as a result of Lost Update or Write Skew
anomalies. However, applications with non-monotonic state mutation
will not, in general, be able to maintain application-level
consistency constraints. Similarly, applications requiring write
visibility between clients should opt for unavailability instead of
possibly missing a write visibility deadline.

\vspace{.5em}\noindent\textbf{TPC-C} Absent a necessary and sufficient result describing exactly which
application logics can be achieved with high availability and to make
these requirements more concrete, we consider an example application:
the TPC-C benchmark. Our goal is \textit{not} to provide a performance
analysis, as any HAT configuration will be non-compliant. Rather, we
examine the five queries contained in the benchmarks---ostensibly
representative of a standard transaction processing application---to
understand which are achievable in a HAT context. TPC-C consists of
five transactions, capturing operation of a real-world warehouse,
including sales, payments, and deliveries. We analyze each kind of
transaction and its interactions below.

\vspace{.5em}\noindent\textit{New-Order} places an order for a
variable quantity of data items, updating warehouse stock as
needed. The transaction selects a sales district, assigns the order an
ID number, adjusts the remaining warehouse stock, and writes a
placeholder entry for the pending order. There are two mutations we
must be consider in detail.

First, assigning an order ID number is challenging. We can easily
assign a \textit{unique} order ID, by, say, concatenating the customer
ID with a timestamp and server ID. However, the TPC-C specification
requires that order numbers are assigned sequentially, which we cannot
guarantee in HATs because assigning sequential IDs requires preventing
Lost Update. Accordingly, HATs can allow placement of new orders with
unique identifiers but cannot provide compliant TPC-C execution due to
the requirement for a non-monotonic counter increase.

Second, decrementing inventory counts is, in general, a potentially
risky operation because, without coordination, inventory counts may
become negative. However, TPC-C automatically ``restocks'' inventory
counts (by 91 items) when there are 10 or fewer items
remaining. Accordingly, even in the presence of concurrent new orders
to the same item, the stock will never fall below zero. However, an
$R$ replica HAT system might end up with $91(R-1)$ more stock than in
a single-site implementation.

\vspace{.5em}\noindent\textit{Payment} updates running balances for
warehouses, districts, and customer records as well as providing an
audit trail. All balance increase operations can be made commutative,
so the transaction can execute under HATs.

\vspace{.5em}\noindent\textit{Order-Status} is a read-only transaction
and can be safely executed under HATs. The TPC-C specification does
not require recency bounds but, in practice, they may be desirable. If
a client is sticky with a set of replicas, she will read her own
writes, which may be sufficient.

\vspace{.5em}\noindent\textit{Delivery} represents the fulfillment of
a pending order created by a New Order transaction: it deletes the
order from the pending list, updates the customer's balance, updates
the order's carrier ID and delivery time, and updates the customer
balance. The challenge in delivery comes in deletion, which is
non-monotonic: an item might be removed twice from the pending
table. If delivery can be made idempotent---that is, if multiple
Delivery transactions have the same effect as one---then multiple
clients can execute the same Deliveries without fatal consequences.

In practice, given that a physical shipment is likely only delivered
once, it is unlikely that multiple agents will attempt a Delivery
transaction for the same order simultaneously. However, in this case,
we have simply moved the non-monotonicity to the real world: the
carrier that picks up the package for an order must ensure that she is
the only carrier who has done so. While we can program around the
other operations in the order, HATs cannot--without the help of
business-specific processes--serve Delivery transactions. That said,
New-Orders may be placed with high availability.

\vspace{.5em}\noindent\textit{Stock-Level} is another read-only
transaction and can be executed without any non-HAT guarantees, with
the same caveats regarding recency.

\vspace{.5em} Throughout execution, TPC-C also requires the
maintainence of several integrity constraints. For example,
Consistency Condition 1 (3.3.2.1) requires that each warehouse's sales
count must reflect the sum of its subordinate sales districts. This
integrity constraint spans two tables but, given the ability to update
both tables atomically via transactional atomicity, can be easily
maintained. Consistency Conditions 4 through 12 (3.3.2.4-12) can
similarly be satisfied by applying updates atomically across
tables. Consistency Conditions 2 and 3 (3.3.2.2-3) concern order ID
assignment and are more problematic, as we have discussed.

In summary, many---but not all---TPC-C transactions are well-served by
HATs. The two problematic transactions---New-Order and Payment---rely
on non-monotonic state updates. The former transaction can be modified
to ensure ID uniqueness at the expense of sequential per-district ID
ordering, while the latter is an inherently non-monotonic action
requiring external compensation or stronger consistency protocols. We
expect that, especially for read-dominated workloads found in many
online services, HAT guarantees may suffice for many
applications. Finally, while TPC-C is not subject to write skew
anomalies, we note that many TPC-E isolation tests, such as one that
requires simultaneously modifying a product description and its image,
are satisfied by HATs.

### 2013-03-26 writeup/futurework.tex
Another pass at writing

@@

The current state of the art places several useful models within
various categories in the availability space. This is still a
practical result---take, for instance, CAP's impact on distributed
systems of the last decade. But what property defines the boundaries
between classes of highly available systems? Is there a meta-property
that exactly characterizes what properties are achievable with high
availability or sticky high availability? This is an area of ongoing
research within the HAT project.

\textbf{What are the costs and benefits within HAT designs?} There are
many possible combinations of HAT guarantees, and the performance and
ease of implementation of a HAT system will likely vary depending on
which guarantees it attempts to provide. As we have alluded to, the
cost in moving from a sticky high available system to a truly highly
available system seems expensive. As a concrete example, the Eiger and
COPS systems use metadata garbage collection and assume linearizable
clients within each datacenter in order to keep the overhead of
providing causal consistency small. Without performing extensive
client-side caching, these techniques do not seem applicable to a
highly available implementation. This is not intended as a fault of
the prior systems---they perform well for the task at hand---but hint
at the complexity of moving between sticky and truly highly available
systems.

\textbf{Where do semantics-based concurrency control lie in the HAT
  landscape?} The vast amount of literature on semantics-based
concurrency control provides hints as to when coordination can be
avoided but still providing ``strongly consistent'' outcomes. We have
performed a narrow case study in this work, but a general
taxonomization of techniques such as escrow transactions and Sagas
could help answer the challenge of programming in a weakly consistent
environment. These techniques might be aided by considering models
with service-level agreements or bounded messaging asynchrony. While
our focus is on ``classic'' distributed systems trade-offs, there are
several interesting hybrids to consider.

>>>>

We believe that there are several promising avenues for future work on
Highly Available Transaction systems:

@@

This paper places several useful models within various categories in
the availability space. This is still a practical result---take, for
instance, CAP's impact on distributed systems of the last decade. But
what property defines the boundaries between classes of highly
available systems? Is there a meta-property that exactly characterizes
what properties are achievable with high availability or sticky high
availability? This is an area of ongoing research within the Berkeley
HAT project.

\textbf{What are the costs and benefits within HAT designs?} There are
many possible combinations of HAT guarantees, and the performance and
ease of implementation of a HAT system will likely vary depending on
which guarantees it attempts to provide. As we have discussed, the
cost in moving from a sticky high available system to a truly highly
available system seems expensive. As a concrete example, the Eiger and
COPS systems~\cite{cops, eiger} use metadata garbage collection and
assume linearizable clients within each datacenter in order to keep
the overhead of providing causal consistency small. Without performing
extensive client-side caching, these techniques do not seem applicable
to a highly available implementation. This is not intended as a fault
of the prior systems---they perform well for the task at hand---but
hint at the complexity of moving between sticky and truly highly
available systems.

\textbf{Where do semantics-based concurrency control lie in the HAT
  landscape?} The vast amount of literature on semantics-based
concurrency control provides hints as to when coordination can be
avoided but still providing ``strongly consistent'' outcomes. We have
performed a narrow case study in this work, but a general
taxonomization of techniques such as escrow transactions~\cite{escrow}
and Sagas~\cite{sagas} could help answer the challenge of programming
in a weakly consistent environment. These techniques might be aided by
considering models with service-level agreements or bounded messaging
asynchrony. While our focus is on ``classic'' distributed systems
trade-offs, there are several interesting hybrids to consider.
